Capabilities, MMIO and Robust Safety

### 2020-03-31 capability-notes/driver-notes/main.tex
Updated notation for (im)pure invariants over traces

\DeclareMathOperator{\initOKo}{init_{OK}}
\DeclareMathOperator{\initOKos}{init^{\X{sem}}_{OK}}
\newcommandx{\initOK}[3][2=r,3=m]{\initOKo(#1,#2,#3)}
\newcommandx{\initOKS}[4][2=r,3=m,4=W]{\initOKos(#1,#2,#3,#4)}
\DeclareMathOperator{\initBCo}{init_{BC}}
\DeclareMathOperator{\initBCos}{init^{\X{sem}}_{BC}}
\newcommandx{\initBC}[2][1=r,2=m]{\initBCo(#1,#2)}
\newcommandx{\initBCS}[3][1=r,2=m,3=W]{\initBCos(#1,#2,#3)}
\DeclareMathOperator{\driverC}{code}
\DeclareMathOperator{\trace}{tr}
\DeclareMathOperator{\lowval}{\Sf{lowval}}
\DeclareMathOperator{\lifto}{\Sf{lift}}
\newcommandx{\trInv}[2][1=P,2=\iota]{\operatorname{trInv}^{#2}(#1)}
\newcommand{\wrap}[1]{\operatorname{wrap}_{\mathrm{#1}}}
\newcommand{\bigast}[2]{\underset{#1}\Sep{\!\!\! #2}\;}
\newcommand{\app}{\mathbin{+\!\!+}}

@@

\newcommand{\Wpub}{\sqsupseteq^{\X{pub}}}
\newcommand{\Wpriv}{\sqsupseteq^{\X{priv}}}
\begin{mathpar}
  \mprset{vskip=0.5em}
  \inferrule
  { \forall a_{i} \in A_{\X{entry}} \ldotp\,
      \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\!
         \vdash
         \forall W \ldotp\;\ER^{\notMMIO} (W) (\X{RX}, g, b, e, a_i)
       \\
     \\
    (r[\overline{r_{a_{i}} := (\X{E}, g, b, e, a_i)}], m, \emptyset) \longrightarrow^*
    (r', m', t)
  }
  {P(t)}
\end{mathpar}

@@

It should be noted that a more general, semantic version of this specification
is also possible. It would contain the following elements, that should all
follow directly from the specification above (TODO: write this out properly and
do the same for the semantic bootspec):
\begin{itemize}
\item All registers except \X{PC} are in the value relation with respect to
  either the empty world $\Wemp$ (if we do not want to allow any non-empty, non-enter
  capabilities in registers), or a world $\Winit$ containing a single standard region for
  all of the adversary's memory.
  The \X{PC} register satisfies both the read- and write conditions (note the
  parallel with the fundamental theorem) with respect to this same world, if we
  picked the non-empty world, or just points to the adversary's memory, which
  does not contain capabilities (if we
  picked the empty world, a capability pointing into memory can never be valid).
  It feels slightly less artificial to pick the non-empty world option here.
\item We have possession of the predicates $\I{region}$ and
  $\I{sts\_full\_world}$ for the world we picked in the previous bullet. we will
  need these to apply the fundamental theorem.
\item A description of the exact driver code, still in terms of points-to chunks
  (or, in this case equivalently, an invariant). We can write this invariant as
      $\driverC(b,e,\overline{\V{op}_l}) = \knowInv{\iota}{\bigast{l \in [b,e)}{l \mapsto_{a} \mathit{\V{w}_l}}}$, with $\V{w}_l$
      some hard-coded driver word (an instruction or a capability) at location $l$.
\end{itemize}

@@

\begin{mathpar}
  \mprset{vskip=0.5em}
  \inferrule
  { \forall a_{i} \in A_{\X{entry}} \ldotp\,\left\{
    {\begin{array}{ll}
    \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\ast
    \driverC(b,e,\overline{\V{w}_l})\!
         \vdash \\
         \forall W \Wpriv \Winit \ldotp\;\ER^{\notMMIO} (W) (\X{RX}, g, b, e, a_i)
        \end{array}} \right. \\
   \initOKS{[b,e)}[r][m][\Winit] \\
  (r[\overline{r_{a_{i}} := (\X{E}, g, b, e, a_i)}], m, \emptyset) \longrightarrow^* (r', m', t)\\
  }
  {P(t)}
\end{mathpar}

@@

\begin{mathpar}
  \mprset{vskip=0.5em}
  \inferrule
  {  \forall a_{i} \in A_{\X{entry}} \ldotp\,\left\{
    {\begin{array}{ll}
    \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\ast
    \driverC(b,e,\overline{\V{op}_l})\!
         \vdash \\
         \forall W \Wpriv \Winit \ldotp\;\ER^{\notMMIO} (W) (\X{RX}, g, b, e, a_i)
        \end{array}} \right. \\   \X{bootSpec}^{\X{sem}}(P,g,b,e,A_{\X{entry}}, \Winit) \\
  (r_0, m_0, \emptyset) \longrightarrow^* (r, m, t)\\
  }
  {P(t)}
\end{mathpar}

@@

We have to tie this integer $n$ to the number of I/O events
present in the trace $t$, otherwise it will prove impossible to verify our
driver. Concretely, the invariant we will verify our driver under changes from
$\knowInv{\iota}{\exists t\ldotp \tracefrag{t} \ast \pure{|t| \le 1000}}$ to
$\knowInv{\iota}{\exists t\ldotp \tracefrag{t} \ast \pure{|t| \le 1000} \ast
  \exists n\ldotp l_n \mapsto n \ast n = |t| }$. Concretely, this means that
instead of having a pure predicate $P$ in our robust safety theorem, we should
also allow non-pure predicates $P_{\X{np}}$ that entail $P$, i.e. $\forall t
\ldotp P_{\X{np}}(t) \wand \pure{P(t)} $, since we might have to enforce
other invariants relating to local state over the trace as well.

>>>>

\DeclareMathOperator{\initOKo}{init_{OK}}
\DeclareMathOperator{\initOKos}{init^{\X{sem}}_{OK}}
\newcommandx{\initOK}[3][2=r,3=m]{\initOKo(#1,#2,#3)}
\newcommandx{\initOKS}[4][2=r,3=m,4=W]{\initOKos(#1,#2,#3,#4)}
\DeclareMathOperator{\initBCo}{init_{BC}}
\DeclareMathOperator{\initBCos}{init^{\X{sem}}_{BC}}
\newcommandx{\initBC}[2][1=r,2=m]{\initBCo(#1,#2)}
\newcommandx{\initBCS}[3][1=r,2=m,3=W]{\initBCos(#1,#2,#3)}
\DeclareMathOperator{\driverC}{code}
\DeclareMathOperator{\trace}{tr}
\DeclareMathOperator{\lowval}{\Sf{lowval}}
\DeclareMathOperator{\lifto}{\Sf{lift}}
\newcommandx{\trInv}[3][1=P,2=\iota,3]{\operatorname{trInv}^{#2}_{#3}(#1)}
\newcommandx{\trInvNp}[2][1=P,2=\iota]{\trInv[#1][#2][\mathrm{np}]}
\newcommand{\wrap}[1]{\operatorname{wrap}_{\mathrm{#1}}}
\newcommand{\bigast}[2]{\underset{#1}\Sep{\!\!\! #2}\;}
\newcommand{\app}{\mathbin{+\!\!+}}

@@

\newcommand{\Wpub}{\sqsupseteq^{\X{pub}}}
\newcommand{\Wpriv}{\sqsupseteq^{\X{priv}}}
\begin{mathpar}
  \mprset{vskip=0.5em}
  \inferrule
  { \forall a_{i} \in A_{\X{entry}} \ldotp\,
      \trInv\!
         \vdash
         \forall W \ldotp\;\ER^{\notMMIO} (W) (\X{RX}, g, b, e, a_i)
       \\
     \\
    (r[\overline{r_{a_{i}} := (\X{E}, g, b, e, a_i)}], m, \emptyset) \longrightarrow^*
    (r', m', t)
  }
  {P(t)}
\end{mathpar}

@@

It should be noted that a more general, semantic version of this specification
is also possible. It would contain the following elements, that should all
follow directly from the specification above (TODO: write this out properly and
do the same for the semantic bootspec):
\begin{itemize}
\item All registers except \X{PC} are in the value relation with respect to
  either the empty world $\Wemp$ (if we do not want to allow any non-empty, non-enter
  capabilities in registers), or a world $\Winit$ containing a single standard region for
  all of the adversary's memory.
  The \X{PC} register satisfies both the read- and write conditions (note the
  parallel with the fundamental theorem) with respect to this same world, if we
  picked the non-empty world, or just points to the adversary's memory, which
  does not contain capabilities (if we
  picked the empty world, a capability pointing into memory can never be valid).
  It feels slightly less artificial to pick the non-empty world option here.
\item We have possession of the predicates $\I{region}$ and
  $\I{sts\_full\_world}$ for the world we picked in the previous bullet. we will
  need these to apply the fundamental theorem.
\item A description of the exact driver code, still in terms of points-to chunks
  (or, in this case equivalently, an invariant). We can write this invariant as
      $\driverC(b,e,\overline{\V{w}_l}) = \knowInv{\iota}{\bigast{l \in [b,e)}{l \mapsto_{a} \mathit{\V{w}_l}}}$, with $\V{w}_l$
      some hard-coded driver word (an instruction or a capability) at location $l$.
\end{itemize}

@@

\begin{mathpar}
  \mprset{vskip=0.5em}
  \inferrule
  { \forall a_{i} \in A_{\X{entry}} \ldotp\,\left\{
    {\begin{array}{ll}
    \trInv \ast
    \driverC(b,e,\overline{\V{w}_l})\!
         \vdash \\
         \forall W \Wpriv \Winit \ldotp\;\ER^{\notMMIO} (W) (\X{RX}, g, b, e, a_i)
        \end{array}} \right. \\
   \initOKS{[b,e)}[r][m][\Winit] \\
  (r[\overline{r_{a_{i}} := (\X{E}, g, b, e, a_i)}], m, \emptyset) \longrightarrow^* (r', m', t)\\
  }
  {P(t)}
\end{mathpar}

@@

\begin{mathpar}
  \mprset{vskip=0.5em}
  \inferrule
  {  \forall a_{i} \in A_{\X{entry}} \ldotp\,\left\{
    {\begin{array}{ll}
    \trInv \ast
    \driverC(b,e,\overline{\V{w}_l})\!
         \vdash \\
         \forall W \Wpriv \Winit \ldotp\;\ER^{\notMMIO} (W) (\X{RX}, g, b, e, a_i)
        \end{array}} \right. \\   \X{bootSpec}^{\X{sem}}(P,g,b,e,A_{\X{entry}}, \Winit) \\
  (r_0, m_0, \emptyset) \longrightarrow^* (r, m, t)\\
  }
  {P(t)}
\end{mathpar}

@@

We have to tie this integer $n$ to the number of I/O events
present in the trace $t$, otherwise it will prove impossible to verify our
driver. Concretely, the invariant we will verify our driver under changes from
$\trInv[|t| \le 1000]$ to
$\trInvNp[\pure{|t| \le 1000} \ast
  \exists n\ldotp l_n \mapsto n \ast n = |t| ]$, where we defined
%
\[
  \begin{array}{lcl}
    \trInvNp[P_{\X{np}}] & \eqdef & \knowInv{\iota}{\exists t.\; \trace(t) \ast P_{\X{np}}(t)}
  \end{array}
\]

(i.e., the predicate over the trace does not need to be pure any longer).
Concretely, this means that
instead of having a pure predicate $P$ in our robust safety theorem, we should
also allow non-pure predicates $P_{\X{np}}$ that entail $P$, i.e. $\forall t
\ldotp P_{\X{np}}(t) \wand \pure{P(t)} $, since we might have to enforce
other invariants relating to local state over the trace as well.
